---
title: "Preprocessing"
author: "Billy Parmenter"
date: "2023-06-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(corrplot)){install.packages("corrplot")}
library(corrplot)

if (!require(caret)) {install.packages("caret")}
library(caret)

if (!require(gbm)) {install.packages("gbm")}
library(gbm)

if (!require(randomForest)) {install.packages("randomForest")}
library(randomForest)
```


# Data Exploration


```{r}
# Load data
RentalData <- read.csv('../Data/House_Rent_Dataset.csv')
head(RentalData)
```

There seems to be no missing data
```{r}
summary(RentalData)
```


# Data Preprocessing
These steps were taken from the slides in week 1


## 1) Deduplicate your data
```{r}
numberOfDuplicate <- length(RentalData) - length(unique(RentalData))
numberOfDuplicate
```
There are no duplicate rows

## 2) Deal with missing data

```{r}
RentalData
RentalData <- na.omit(RentalData)
RentalData
```

## 3) Fix structural errors

Floor as it is does not seem usable. Try spliting the floor feature into three other numeric features, Rental.Floor (the floor the rental unit is on), Building.Floors (How many floors are in the building), and Rental.Floor.Ratio (Rental.Floor/Building.Floors, how far up the unit is to the top of the building)

Getting the number of floors in a building
```{r}
if(!require(stringr)){install.packages("stringr")}
library(stringr)
# Use regex to get the top floor number and convert character to numeric
RentalData$Building.Floors <- str_extract(RentalData$Floor, "(?<=out of )(\\d)")
RentalData$Building.Floors <- as.numeric(RentalData$Building.Floors)

# Add one because they use ground floor as 1st floor and 1st floor as 2nd floor...
RentalData$Building.Floors <- RentalData$Building.Floors + 1
```

Getting the floor of the rental unit
```{r}
# Get all numeric values for the rental units floor
RentalData$Rental.Floor <- str_extract(RentalData$Floor, "^(\\d)")
RentalData$Rental.Floor <- as.numeric(RentalData$Rental.Floor)

# Add one because they use ground floor as 1st floor and 1st floor as 2nd floor...
RentalData$Rental.Floor <- RentalData$Rental.Floor + 1

# Get all values of ground floor units and set it as floor 1
RentalData$Rental.Floor[str_detect(RentalData$Floor, "^Ground")] <- 1 
```

There are some units that are listed as upper/lower basements I am setting these to 0
```{r}
# Set any basement units to floor 0
RentalData$Rental.Floor[str_detect(RentalData$Floor, "^Lower|Upper")] <- 0 
```

Getting the ratio of unit floor to building floors
```{r}
RentalData$Rental.Floor.Ratio <- RentalData$Rental.Floor / RentalData$Building.Floors * 100
```

By splitting the data we have created 4 NA values
```{r}
summary(RentalData)
```


## 4) Remove irrelevant data

Look at correlation and see if there are any features that are not relevant to the problem
```{r}
noNaRentalData <- na.omit(RentalData)
res <- cor(noNaRentalData[,c(2,3,4,11,13,14,15)],method="pearson")

corrplot(res, type = "upper", diag = FALSE,
         tl.col = "black", tl.srt = 45)
```
There seems to be little correlation between floor the rental unit is on and rent as well as ratio and rent but no correlation between number of floors in the building and rent.
```{r}
summary(RentalData$Rental.Floor)
summary(RentalData$Building.Floors)
# Find rows with missing data in RentalData
missing_rows <- which(rowSums(is.na(RentalData)) > 0)

# Display the rows with missing data
RentalData[missing_rows, ]
```

## 5) Filter out features of Building.Floors
```{r}
RentalData
RentalData <- na.omit(RentalData)
RentalData

col_to_drop <- c('Building.Floors', 'floor', 'Posted.On', 'Area.Type', 'Area.Locality', 'Tenant.Preferred', 'Point.of.Contact')

RentalData <- RentalData[, !(names(RentalData) %in% col_to_drop)]
RentalData
res <- cor(RentalData[,c(1,2,3,7,8,9)],method="pearson")

corrplot(res)
```


## 6) Filter out data outliers
```{r}

create_density_plots <- function(dataframe) {
  numeric_columns <- sapply(dataframe, is.numeric)
  
  for (col in colnames(dataframe)[numeric_columns]) {
    p <- ggplot(dataframe, aes(x = dataframe[[col]])) +
      geom_density(fill = "blue", alpha = 0.5) +
      labs(title = paste("Density Plot of", col))
    
    print(p)
  }
}

create_box_plots <- function(dataframe) {
  numeric_columns <- sapply(RentalData, is.numeric) # Identify numeric columns

  for (col in colnames(RentalData)[numeric_columns]) {
    boxplot(RentalData[[col]], horizontal=TRUE, main=col)
  }
}

create_density_plots(RentalData)
create_box_plots(RentalData)


```
Deal with Size outliers
```{r}

remove_outliers <- function(dataframe) {
  numeric_columns <- sapply(dataframe, is.numeric)
  
  for (col in colnames(dataframe)[numeric_columns]) {
    if (col != 'Rental.Floor'){
      outliers <- boxplot.stats(dataframe[[col]])$out # Get outliers
      dataframe[[col]][dataframe[[col]] %in% outliers] <- NA
    }
  }
  cleaned_dataframe <- na.omit(dataframe)
  return(cleaned_dataframe)
}


RentalData_cleaned <- remove_outliers(RentalData)

create_density_plots(RentalData_cleaned)
create_box_plots(RentalData_cleaned)

```
The number of removed rows
```{r}
dim(RentalData)[1] - dim(RentalData_cleaned)[1]
```
```{r}

label_encode_all_columns <- function(dataframe) {
  for (col in colnames(dataframe)) {
    if (is.factor(dataframe[[col]]) || is.character(dataframe[[col]])) {
      dataframe[[col]] <- as.numeric(factor(dataframe[[col]]))
    }
  }
  return(dataframe)
}

RentalData_encoded <- label_encode_all_columns(RentalData_cleaned)
RentalData_encoded

```

## 7) Data Visulization
```{r}
# pie chart for city
city_freq <- table(RentalData_encoded$City)
city_perc <- round(100*city_freq / sum(city_freq), 1)
labels <- paste(names(city_freq), "(", city_perc, "%)")
pie(city_freq, labels=labels, main='City')

# plot rent
barplot(table(RentalData_encoded$Rent), cex.names=.75, main='Rent')


```



```{r}
pairs(RentalData_encoded)
```

```{r}
my_cols <- c("#00AFBB", "#E7B800", "#FC4E07")  
pairs(RentalData_encoded, pch = 19,  cex = 0.5)

# Correlation panel
panel.cor <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y, pch = 19)
}
# Create the plots
pairs(RentalData_encoded, 
      lower.panel = panel.cor,
      upper.panel = upper.panel)
```

```{r}
library(psych)
pairs.panels(RentalData_encoded, 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
```
The city that the unit is in plays a big role, maybe try one hot encoding. Also Furnishing is a factor
```{r}
aggregate(RentalData$Rent, list(RentalData$City), FUN=mean)
aggregate(RentalData$Rent, list(RentalData$Furnishing.Status), FUN=mean)
```

```{r}
train_model <- function(data, method, target_col) {
  train_index <- createDataPartition(data[[target_col]], p = 0.8, list = FALSE)
  train_data <- data[train_index, ]
  
  trained_model <- train(
    formula(paste(target_col, "~ .")),
    data = train_data,
    method = method,
    trControl = trainControl(method = "cv"),
    preProcess = c("center", "scale"),
    verbose = FALSE
  )

  return(trained_model)
}

train_model_stepwise <- function(data, target_col) {
  train_index <- createDataPartition(data[[target_col]], p = 0.8, list = FALSE)
  train_data <- data[train_index, ]
  
  formula_initial <- paste(target_col, "~ .")
  
  # Train a full linear regression model
  full_lm_model <- train(
    as.formula(formula_initial),
    data = train_data,
    method = "lm",
    trControl = trainControl(method = "cv"),
    preProcess = c("center", "scale")
  )
  
  print("Full Model:")
  print(summary(full_lm_model$finalModel))
  
  # Perform backward selection
  backward_formula <- step(
    lm(as.formula(formula_initial), data = train_data),
    direction = "backward"
  )
  
  print("Backward Selection Model:")
  print(summary(lm(backward_formula, data = train_data)))
  
  # Train a linear regression model with the selected features from backward selection
  best_lm_model <- train(
    as.formula(backward_formula),
    data = train_data,
    method = "lm",
    trControl = trainControl(method = "cv"),
    preProcess = c("center", "scale")
  )
  
  print("Best Model:")
  print(summary(best_lm_model$finalModel))
  
  return(best_lm_model)
}

# Train a Linear Regression model using stepwise selection
lm_model <- train_model_stepwise(RentalData_encoded, target_col = "Rent")



# Train a Random Forest model
rf_model <- train_model(RentalData_encoded, method = "rf", target_col = "Rent")

# Train a Gradient Boosting model
gbm_model <- train_model(RentalData_encoded, method = "gbm", target_col = "Rent")
```


```{r}


# Function to evaluate a trained model
evaluate_model <- function(trained_model, test_data, target_col, model_name) {
  predictions <- predict(trained_model, newdata = test_data)
  residuals <- test_data[[target_col]] - predictions
  rmse <- sqrt(mean(residuals^2))
  
  # Calculate R-squared
  ss_total <- sum((test_data[[target_col]] - mean(test_data[[target_col]]))^2)
  ss_residual <- sum(residuals^2)
  r_squared <- 1 - (ss_residual / ss_total)
  
  # Calculate adjusted R-squared
  n <- nrow(test_data)
  p <- length(trained_model$finalModel$coefficients)
  adjusted_r_squared <- 1 - ((1 - r_squared) * (n - 1)) / (n - p - 1)
  
  # Print metrics
  cat("Root Mean Squared Error (RMSE):", rmse, "\n")
  cat("R-squared:", r_squared, "\n")
  cat("Adjusted R-squared:", adjusted_r_squared, "\n")
  
  # Plot residuals vs fitted values
  plot_residuals <- plot(residuals, 
                          xlab = "Fitted Values", ylab = "Residuals",
                          main = paste("Residuals vs Fitted Values ", model_name))
  abline(h=0, col="red")
  print(plot_residuals)
  
  # Plot residuals histogram
  plot_histogram <- qplot(residuals, bins = 20, 
                           xlab = "Residuals", ylab = "Frequency",
                           main = paste("Histogram of Residuals ", model_name))
  print(plot_histogram)
  
  # Plot normal QQ plot of residuals
  plot_qq <- qqnorm(residuals, main = paste("Normal Q-Q Plot of Residuals ", model_name))
  qqline(residuals)
  print(plot_qq)
  
  return(list(rmse = rmse, r_squared = r_squared, adjusted_r_squared = adjusted_r_squared))
}

# Evaluate the three models
evaluation_lm <- evaluate_model(lm_model_stepwise, RentalData_encoded, target_col = "Rent", "Linear")
evaluation_rf <- evaluate_model(rf_model, RentalData_encoded, target_col = "Rent", "Random Forest")
evaluation_gbm <- evaluate_model(gbm_model, RentalData_encoded, target_col = "Rent", "GBM")

# Compare the metrics
comparison <- data.frame(
  Model = c("Linear Regression", "Random Forest", "Gradient Boosting"),
  RMSE = c(evaluation_lm$rmse, evaluation_rf$rmse, evaluation_gbm$rmse),
  R2 = c(evaluation_lm$r_squared, evaluation_rf$r_squared, evaluation_gbm$r_squared),
  Adjusted_R2 = c(evaluation_lm$adjusted_r_squared, evaluation_rf$adjusted_r_squared, evaluation_gbm$adjusted_r_squared)
)

comparison




```
The Random Forest model seems to have the best performance with lowest RMSE.
The residual plot of random forest model seems pretty close to 0 which is good.


## 8) Prediction
```{r}
# create prediction dataset manually
prediction_data <- data.frame(
  BHK = c(2, 2, 2, 2, 2, 2, 2),
  Rent = c(10000, 20000, 17000, 10000, 7500, 7000, 10000),
  Size = c(1100, 800, 1000, 800, 850, 600, 700),
  Floor = c(213, 14, 14, 10, 10, 207, 216),
  City = rep(5, 7), # Assuming 5 is the encoded value for Kolkata
  `Furnishing.Status` = c(3, 2, 2, 3, 3, 3, 3),
  Bathroom = c(2, 1, 1, 2, 2, 2, 2),
  `Rental.Floor` = c(1, 2, 2, 1, 1, 1, 1),
  `Rental.Floor.Ratio` = c(33.33333, 50.00000, 50.00000, 33.33333, 33.33333, 33.33333, 33.33333)
)
```

```{r}
lm_predictions <- predict(lm_model, newdata = prediction_data)
rf_predictions <- predict(rf_model, newdata = prediction_data)
gbm_predictions <- predict(gbm_model, newdata = prediction_data)

# Display predictions for each model
cat("Linear Regression Predictions:", lm_predictions, "\n")
cat("Random Forest Predictions:", rf_predictions, "\n")
cat("Gradient Boosting Predictions:", gbm_predictions, "\n")
```

